Data Specialization for a Staged Lambda Calculus

### 2015-06-24 paper/semantics.tex

We express two-stage programs as terms in \lang, a typed, modal lambda calculus. 
The type system of \lang\ (\ref{sec:statics}) classifies both the stage and type of a 
computation, and is sufficient to show that the staged dynamics of \lang\ (\ref{sec:dynamics})
do not get stuck.

@@

The typing judgment $\typeswor e A$, defined in \ref{fig:statics}, means that
$e$ has type $A$ at world $w$, in the context $\Gamma$. 
Here, a \emph{world} is a slightly finer classification than a stage.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonep\ for \emph{ground} first-stage computations, and 
\bbonem\ for first-stage computations containing a \emph{mix} of code from each stage. 
The distinction between these two first-stage worlds is necessitated by our splitting algorithm,
and will be discussed in \ref{sec:needGround}.

@@

\subsection{(Small Step) Dynamics}
\label{sec:dynamics}

We provide a small-step semantics which,
given a \lang\ term $e$, evaluates its first stage subterms
either until $e$ is a value (if $e$ is at world \bbonep\ or \bbonem)
or until all of $e$'s first stage subterms have been evaluated away
(when $e$ is at world \bbtwo).
Because first- and second-stage code are
interleaved in \lang\ terms, this first pass of
evaluation requires some care.

Moreover, our staged semantics contains one innovation over those of 
\cite{davies96} and \cite{taha-thesis-99}, which is the addition
of {\em hoisting steps}.  
To illustrate, consider the example
\begin{lstlisting}
prev{let x = next{2*3} in next{prev{x}+4}}
\end{lstlisting}
A non-staged semantics which ignores $\next$s and $\prev$s
would proceed on this term by reducing \verb|2*3| to \verb|6| in place.
A staged semantics does not, because it only reduces first stage redexes, and
\verb|2*3| is at the second stage.  
The staged semantics of \cite{davies96} and \cite{taha-thesis-99} instead proceed by
substituting \verb|next{2*3}| in for \verb|x| to yield
\begin{lstlisting}
prev{next{prev{next{2*3}}+4}}
\end{lstlisting}
This takes one step to 
\begin{lstlisting}
prev{next{2*3+4}}
\end{lstlisting}
and then one more step to
\begin{lstlisting}
2*3+4
\end{lstlisting}
which is the final entirely-second-stage result, called a {\em residual}.
Our semantics differs from this formulation in one way.
Consider a slightly modified example where \verb|x| appears twice:
\begin{lstlisting}
prev{let x = next{2*3} in next{prev{x}+prev{x}}}
\end{lstlisting}
A standard stage semantics would, in four steps, evaluate this to 
\begin{lstlisting}
2*3+2*3
\end{lstlisting}
Note how the multiplication is duplicated in the residual.
We find this behavior potentially undesirable for our applications,
so our semantics instead {\em hoists} the contents of a next statement 
to the outermost containing $\prev$, turning
\begin{lstlisting}
prev{let x = next{2*3} in next{prev{x}+prev{x}}}
\end{lstlisting}
into
\begin{lstlisting}
let y = 2*3 in
prev{let x = next{y} in next{prev{x}+prev{x}}}
\end{lstlisting}
This then evaluates to 
\begin{lstlisting}
let y = 2*3 in y+y
\end{lstlisting}
in four steps.  
This hoisting step occurs as the last step in the evaluation of a $\next$.

Formally, the dynamic semantics comprises three judgements: 
steps $\stepwor e e'$, which asserts that a term $e$ performs some internal reduction to produce $e'$,
where both terms are at world $w$;
doneness $e \done$, which asserts that a term $e$ at world $w$ has no more first stage redexes;
and hoisting steps $\lift e {e'}$, which are available only at world \bbonem.

One important feature of our semantics is that it reduces {\em all} first stage redexes
in the first pass, even if they occur underneath second stage binders.
This does not cause evaluation to ``get stuck'' however, 
because first stage code cannot depend on the results of second stage code,
as enforced by our type system.

More formally, the type safety of our dynamics can be stated
with the usual progress and preservation lemmas, annotated with a world:
\\{\bf Progress.~} If $\typeswor e A$, then either 
$\stepwor e {e'}$ or $e \done$ or $\lift e {e'}$ (hoisting available only at $w = \bbonem$).
\\{\bf Preservation.~} For normal steps, if $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$. 
For hoisting steps, if $\typesone e A$ and $\lift e {e'}$, then $\typesone {e'} A$.

@@

\paragraph{Done.} The judgement $e \done$ indicates that a term $e$ contains no first stage redexes.
It generalizes over the notions of values and residuals.

\paragraph{Standard Steps.} The step rules are split into reduction rules
which encode the elimination of redexes, 
and substep rules which define where such a reduction can occur within a larger term.
In our semantics, the reduction rules are restricted to worlds \bbonep\ and \bbonem,
so that only first stage redexes are eliminated.
Substep rules are provided for all worlds, as well as for $\next$ and $\prev$.
Also note the extra substep rules at world \bbtwo, which allow steps under binders.

\paragraph{Hoisting Steps.} 
Once the contents of a next block have been reduced to a residual,
they are replaced by a fresh variable and then bound to that
variable in a let term just above the innermost containing prev.
This is acomplished by a hoisting step,
most of whose rules are substep rules which ``bubble up'' the
terms to be bound.

\subsection{(Big Step) Dynamics}
\label{sec:stagedsemantics}

\input{figures/dynamics-overview}

\input{figures/dynamics}

Although the type system of \lang\ (specifically, the opaque nature of $\fut$
types) ensures that we can perform all first-stage computations before any
second-stage ones, the fact that first- and second-stage computations are
interleaved in \lang\ terms means that realizing this notion of staged
evaluation requires some care.

We evaluate \lang\ terms in two stages. The first stage of evaluation has three
mutually recursive modes of operation, depending on the world of the term it is
evaluating. Specifically, first-stage evaluation:
\begin{itemize}
\item fully reduces terms in \bbonep, as they are purely first-stage;
\item reduces terms in \bbonem, save for any world \bbtwo\ terms they may
contain (inside \texttt{next}s);
\item scans terms in \bbtwo\ for any world \bbonem\ subterms (inside \texttt{prev}s), 
and reduces those in place.
\end{itemize}
These tasks are accomplished by the $\redsym$, $\redonesym$, and $\redtwosym$
big-step judgments, respectively (defined in \ref{fig:diaSemantics}).

First-stage evaluation ultimately produces a monostaged term, called a \emph{residual}
(\ref{fig:values}), all of whose redexes are second-stage computations. The
second stage of evaluation is therefore just ordinary (unstaged) evaluation of
this residual, which we notate $\redsym$. (For simplicity, we treat residuals
and \bbonep\ terms as living in a single language equipped with a standard
evaluation $\redsym$, whose definition we omit for space reasons.)

Notice that our type system makes it possible to evaluate terms in a staged
fashion, but doing so still requires some machinery, since the stages are
syntactically interleaved within an individual term. For example,
\texttt{qss} contains a first-stage recursive call inside a second-stage
\texttt{case}; thus, an ordinary outermost-first evaluation strategy would not
be properly staged.

\paragraph{First-stage evaluation at \bbonem\ ($\redonesym$).}
The difficulty of evaluating terms at \bbonem\ is that they contain both terms
at \bbonep\ (via $\curr$) and at \bbtwo\ (via $\fut$); the other evaluation
rules are ordinary call-by-value rules. When we encounter $\pure{e}$, we
evaluate $e$ as a monostaged term; \texttt{letg} and \texttt{caseg} simply reach
inside $\pure$ and evaluate like \texttt{let} and \texttt{case}.

$\next{e}$ is trickier, because we should not evaluate $e$ (at \bbtwo) yet.
There are two difficulties---$e$ may contain other code at \bbonem\ (which we
handle with $\redtwosym$, discussed below); and $\next{e}$ may be part of a
first-stage redex, which we \emph{do} need to reduce. For example, in
\begin{lstlisting} 
1`(fn x : $int => e') (next{`2`e`1`})`
\end{lstlisting}
we must reduce this first-stage function application, but cannot reduce $e$ yet.
We could reduce this to $[\next{e}/x]e'$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to hoist $e$ outside, binding it
to a temporary variable $\hat{y}$, and substituting that variable instead:
\begin{lstlisting} 
2`let val yhat = e in `1`[next{`2`yhat`1`}/x]e'`
\end{lstlisting}
then proceed by evaluating $[\next{\hat{y}}/x]e'$.

We achieve this by threading through a \emph{residual table} $\xi$, which
functions as a top-level sequence of \texttt{let}-bindings of residuals $e$ to
variables $\hat{y}$ in an evaluated \bbonem\ term. Thus, $\redonesym$ evaluation
yields a pair $\rtab \xi v$ of a residual table and a \emph{partial value} $v$
containing variables bound in that table. The grammar of residual tables and
partial values (terms fully evaluated at the first but not second stage) is
given in \ref{fig:values}.

%The input to $\redonesym$ may be open on stage \bbtwo\ variables, but the type
%system ensures that those must occur under a $\next$. Consequently $\redonesym$
%will never directly encounter a variable.

\paragraph{First-stage evaluation at \bbtwo\ ($\redtwosym$).}
Since terms at world \bbtwo\ represent second-stage computations, we should not
evaluate them yet. However, we do need to evaluate any first-stage computations
they contain (inside $\prev$), \emph{even if} those computations are guarded or
underneath binders.

For all constructs other than $\prev{e}$, $\redtwosym$ evaluation proceeds
recursively on all subterms, yielding a monostaged residual $q$. At $\prev{e}$,
we evaluate $e$ using $\redonesym$ to $\rtab \xi v$. Because $\colmix e{\fut A}$,
$v$ must have the canonical form $\next{\hat{y}}$ for some $\hat{y}$ bound in $\xi$. We \emph{reify}
$\xi$ into a sequence of second-stage \texttt{let} bindings around $\hat{y}$,
using the auxiliary judgment $\reify\xi q{q'}$ defined in
\ref{fig:diaSemantics}, then return the resulting residual $q'$.

%Note that partial values and residuals are both terms for which stage \bbone\
%computation has completed; however, partial values are stage~\bbone\ terms,
%while residuals are stage~\bbtwo\ terms.

%As an optimization, we can include the special-case rule,
%\begin{mathpar}
%\inferdiaone [hat] {\red {\next~\hat y}{\cdot,\next~\hat y}}{\cdot}
%\end{mathpar}
%to avoid one-for-one variable bindings in the residual.

>>>>

We express two-stage programs as terms in \lang, a typed, modal lambda calculus.

One of the central concepts in the formalization of \lang\ is the notion of a \emph{world},
which is a slightly finer classification than a stage.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonem\ for \emph{mixed} first-stage computations that may contain second stage subterms with $\next$ blocks,
and \bbonep\ for \emph{ground} first-stage computations that may not.
The distinction between these two first-stage worlds is necessitated by our splitting algorithm,
and will be discussed in \ref{sec:needGround}.

The abstract syntax of \lang\ is presented as a grammar in \ref{fig:grammar}.  
Note that some production rules are parameterized by world.
Our syntax also distinguishes values from general terms [need ``series expression'' citation!], 
as this will be useful in defining our splitting algorithm.
There are three forms of value: \emph{partial values} for \bbonem, \emph{ground values} for \bbonep, and \emph{residuals} for \bbtwo.
All represent computations which have no remaining work in the first stage.

@@

The typing judgment $\typeswor e A$, defined in \ref{fig:statics}, means that
$e$ has type $A$ at world $w$, in the context $\Gamma$.

@@

\subsection{Dynamics}
\label{sec:dynamics}

We provide a small-step semantics which evaluates the first stage subterms
of any \lang\ term.  
This requires notions of values and steps that are specialized to each world.

\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement $\steptwo e {e'}$.
Since no reduction of stage two terms is to take place, it does nothing but search for \prev\ blocks within $e$.
A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
To be a residual, a term must have no first stage subterms (equivalently, no $\prev$s), even within the body of a function or branches of a case.
This implies that $\stepsym \bbtwo$ must proceed underneath second stage binders.

\paragraph{World \bbonep.} Since the ground fragment of the language is not dependent on other worlds, the semantics of ground is just that of a monostage language.
Thus, $\steppure e {e'}$ searches for redexes and reduces them in the standard way, and a ground value $u$ comprises only units, injections, tuples, and functions, where the body of the function may be any ground term.

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, searches for redexes and reduces them in place.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other kinds of steps there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable,
as they would be in the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
For example, consider:
\begin{lstlisting} 
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $\hat{y}$, and substituting that variable instead:
\begin{lstlisting} 
2`let val yhat = e in prev{`1`[next{`2`yhat`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a hoisting step.
They steps are introduced when a \next\ block has contents that are a residual but not already a variable.
In essence, they operate by ``bubbling up'' a subtitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.

\subsection{Type Safety}

The statics and dynamics of \lang\ are related by a type safety theorem
comprising the usual progress and preservation lemmas, again annotated by world:
\\{\bf Progress.~} If $\typeswor e A$, then either 
$\stepwor e {e'}$ or $e \done$ or $\lift e {e'}$ (hoisting available only at $w = \bbonem$).
\\{\bf Preservation.~} For normal steps, if $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$. 
For hoisting steps, if $\typesone e A$ and $\lift e {e'}$, then $\typesone {e'} A$.

@@
